The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-03-13 paper/iphas-dr2.tex
Improves Section 7

A primary motivation for carrying
out the survey 
was to enable the discovery of 
new emission-line objects across the Galactic Plane.
\ha\ in emission is a well-known tracer
for stars in the short-lived pre- or
post-main sequence stages of their evolution,
and hence IPHAS aims to allow larger, deeper
and more statistically robust samples of such rare objects
to be established.
The survey-wide identification and analysis 
of such stars is beyond the scope of the present work,
but in this section we demonstrate how the
catalogue may be used for this purpose.

@@

The target of our demonstration is Sh 2-82:
a 5~arcmin-wide H{\sc ii} region located near $(l,b)=(53.55^\circ, 0.00^\circ)$
in the constellation of Sagitta.
Nicknamed by amateur astronomers as the ``Little Cocoon Nebula'',
Sh 2-82 is ionised by 
the $\sim$10th magnitude star HD\,231616
% Georgelin1973 claims B0III, others claim B0V
with spectral type B0V/III
\citep{Georgelin1973,Mayer1973,Hunter1990}.
The ionising star has been placed at a likely distance of 1.5-1.7 kpc
based on the photometric parallax
\citep{Mayer1973,Lahulla1985,Hunter1990}.

Figure~\ref{fig:mosaic_iphas} shows a 20-by-15 arcmin
colour mosaic centred on Sh 2-82,
composed of our \ha\ (red channel),
$r$ (green channel),
and $i$ (blue channel) images.
The ionising star can be seen as the bright object
near the centre of the H{\sc ii} region,
which is surrounded by a faint reflection nebula
and several dark cloud filaments.
For comparison, Figure~\ref{fig:mosaic_spitzer} shows
a mosaic of the identical region 
as seen by the Spitzer Space Telescope
in the mid-infrared. The Spitzer image
reveals a bubble-shaped structure of warm dust
which surrounds the entire H{\sc ii} region.

@@

Figure~\ref{fig:emitters} presents
the IPHAS colour-colour diagram for the region covered by the
mosaic images.
Gray circles show all objects in the region
which are brighter than $r<20$
and have been flagged as \emph{reliable}
in our catalogue.
The diagram also shows the unreddened main sequence (solid line)
and the expected position of unredded main-sequence stars
with \ha\ in emission at a strength of EW=$-10{\rm \AA}$ (dashed line),
taken from the colour simulations due to \citet{Barentsen2011a}.
Six stars are found to lie confidently above the 
dashed line at the level of $3\sigma$ 
(i.e. the distance between the dashed line is larger than
three times the uncertainty on the $r$-\ha\ colour).
These reliable candidate \ha-emitters
are marked by red triangles in the colour-colour diagram
and their photometry is detailed in Appendix~\ref{app:emitters}.

@@

The spatial distribution of our six candidate emission-line objects
is marked by yellow triangles in the colour mosaic (Fig.~\ref{fig:mosaic_iphas}).
They are likely to be genuine young stars for two reasons.
Firstly, two of our candidates have recently been identified as likely
candidate Young Stellar Objects (YSO)
in an investigation of the region by \cite{Yu2012}.
In their study, the authors used 2MASS and Spitzer data
to select likely young stars by selecting objects with
circumstellar disks based on the infrared colour excess.
Secondly, we find that the four remaining objects are 
also detected in the Spitzer-based image,
although their colours are less extreme
than those identified as likely YSOs by \citeauthor{Yu2012}.

Prior to IPHAS, this region was essentially unstudied
at faint magnitudes in visible light.
\cite{Lahulla1985} reported magnitudes for 8 stars in the optical at 
$V < 15$. In contrast, the IPHAS catalogue offers photometry
for NN stars in the region down to $r<20$.
This demonstrates the ability of IPHAS to providing complimentary
for the wealth of poorly-studied star-forming regions
at low Galactic latitudes,
which have been unveiled in recent years
by the observed wealth of star-forming ``bubbles''
at mid-infrared wave%\cite{Bica2003} was the first to identify a
%near-infrared cluster
%associated with this region using near-infrared 2MASS data,
%and only recently has this cluster been confirmed
%in a short investigation which combined 2MASS
%and Spitzer data \cite{Yu2012}.
%Many of the young stars identified by \citeauthor{Yu2012}
%can be seen as pink and red stars in Fig.~\ref{fig:mosaic_spitzer}.lengths \cite[e.g.][]{Churchwell2006,Simpson2012}.
%In fact the cluster of young stars associated with Sh 2-82
%was only first reported by \cite{Bica2003},
%remained unstudied until the study by \cite{Yu2012}.

%Amongst the six emitters, N are detected at SNR X
%in the Spitzer X micron band.

%This demonstrates how IPHAS is providing complimentary
%data for some of the thousands of star-forming ``bubbles'' have been unveiled at mid-infrared wavelengths \cite[e.g.][]{Churchwell2006,Simpson2012}.
%Optically-unveiled members of such star-forming regions
%may help to constrain extinctions and distances towards such regions \citep{Barentsen2013}.
%Moreover, in \cite{Barentsen2011a} we showed
%how the history of star formation in a region,
%and the possible triggered nature, may be discussed.
%This is particularly true

%TODO: annotate -10 EW text on plot?
%TODO: mention Gaia, e.g. IPHAS for transient identification?

%\subsection{Caveats and lessons learnt}
%\cite{Bica2003} was the first to identify a
%near-infrared cluster
%associated with this region using near-infrared 2MASS data,
%and only recently has this cluster been confirmed
%in a short investigation which combined 2MASS
%and Spitzer data \cite{Yu2012}.
%Many of the young stars identified by \citeauthor{Yu2012}
%can be seen as pink and red stars in Fig.~\ref{fig:mosaic_spitzer}.

@@

The observations included in this release
were found to achieve a median seeing of 1.1 arcsec
and a $5\sigma-$depth of $r=21.2\pm 0.5$, $i=20.0\pm 0.3$, \ha$=20.3\pm 0.3$.
The global calibration and photometric repeatability
is accurate at the level of $\sigma=0.03$ mag,
providing a significant improvement over the 
previous data release.
The source catalogue provides the best-available
single-epoch astrometry and photometry
for 219~million unique sources.

The data-taking strategy developed for IPHAS
have since been reapplied to carry out a companion INT/WFC survey called UVEX
in U/$g$/$r$ \citep{Groot2009},
and also southern counterpart to IPHAS and UVEX 
is being carried out in $u$/$g$/$r$/$i$/\ha\ 
called VPHAS+ (Drew et al, in press).
We hope to re-use the experience gained by this data release
to create similar releases for these companion surveys.

In future work, we aim to draw upon the PanSTARRS photometric
survey to further improve the accuracy of our global calibration.
We will also aim to correct the photometry for the
radial field distortions.
%We may also produce a catalogue which details all detections.

%TODO: mention Gaia?

@@

\section{Candidate emission-line objects towards Sh 2-82}
\label{app:emitters}
\begin{table}
    \begin{tabular}{lccc}
    \toprule
    Name & $r$ & $i$ & \ha  \\
    \midrule
IPHAS2 J192954.40+181026.1& $17.69\pm0.01$ & $16.12\pm0.01$ & $16.19\pm0.01$ \\
IPHAS2 J193011.01+182051.2& $18.55\pm0.02$ & $16.95\pm0.02$ & $17.31\pm0.02$ \\
IPHAS2 J193021.52+181954.5& $19.72\pm0.05$ & $17.94\pm0.03$ & $18.47\pm0.04$ \\
IPHAS2 J193024.45+181938.3& $19.31\pm0.04$ & $17.55\pm0.02$ & $17.99\pm0.03$ \\
IPHAS2 J193033.00+181609.3& $18.25\pm0.01$ & $16.91\pm0.01$ & $16.92\pm0.01$ \\
IPHAS2 J193042.48+182317.4& $19.96\pm0.03$ & $18.11\pm0.03$ & $18.48\pm0.03$ \\
    \bottomrule
    \end{tabular}
    \caption{Candidate \ha-emitters towards Sh 2-82.}
    \label{tbl:emitters}
\end{table}

>>>>

A primary motivation for carrying out the survey 
was to enable the discovery 
of new emission-line objects
across the Galactic Plane.
\ha\ in emission is a well-known tracer
for stars in the short-lived pre- or
post-main sequence stages of their evolution,
and hence IPHAS allows larger, deeper
and more statistically robust samples
of such rare objects to be established.
The survey-wide identification and analysis 
of emission-line objects is beyond the scope
of the present work,
but in this section we demonstrate how the
catalogue may be used for this purpose.

@@

The target of our demonstration is Sh 2-82:
a 5~arcmin-wide H{\sc ii} region located near $(l,b)=(53.55^\circ, 0.00^\circ)$
in the constellation of Sagitta.
Nicknamed by amateur astronomers as the ``Little Cocoon Nebula'',
Sh 2-82 is ionised by 
the $\sim$10th magnitude star HD\,231616
% Georgelin1973 claims B0III, others claim B0V
with spectral type B0V/III
\citep{Georgelin1973,Mayer1973,Hunter1990}.
The ionising star has been placed 
at a likely distance of 1.5-1.7 kpc
based on its photometric parallax
\citep{Mayer1973,Lahulla1985,Hunter1990}.

Figure~\ref{fig:mosaic_iphas} shows a 20-by-15 arcmin
colour mosaic centred on Sh 2-82,
composed of our \ha\ (red channel),
$r$ (green channel),
and $i$ (blue channel) images.
The ionising star can be seen as the brightest object
in the centre of the H{\sc ii} region,
which is surrounded by a faint reflection nebula
and several dark cloud filaments.
For comparison, Figure~\ref{fig:mosaic_spitzer} shows
a mosaic of the same region 
as seen in the mid-infrared by the Spitzer Space Telescope
The infrared image reveals a bubble-shaped structure
of warm dust which surrounds the entire H{\sc ii} region,
and is thought to contain a wealth of infrared-bright
young stellar objects \citep{Yu2012}.
Many of these young objects 
appear in Fig.~\ref{fig:mosaic_spitzer}
as red- and pink-coloured objects,
located predominantly in the north-western part
of the bubble-shaped structure.

@@

Figure~\ref{fig:emitters} presents
the IPHAS colour-colour diagram for 
the 20-by-15 arcmin region shown in the mosaics.
Gray circles show all objects
which are brighter than $r<20$
and have been flagged as \emph{reliable}
in our catalogue.
The diagram also shows the unreddened main sequence (solid line)
and the expected position of unredded main-sequence stars
with \ha\ in emission
at a strength of EW=$-10{\rm \AA}$ (dashed line),
taken from the colour simulations due to \citet{Barentsen2011a}.
Six stars are found to lie confidently above the 
dashed line at the level of $3\sigma$,
that is, the distance between the objects and the dashed line
is larger than three times the uncertainty on the $r$-\ha\ colour.
These reliable candidate \ha-emitters
are marked by red triangles in the colour-colour diagram
and their photometry is detailed in Table~\ref{tbl:emitters}.
The spatial distribution of the six candidate emission-line objects
is marked by yellow triangles in the colour mosaic (Fig.~\ref{fig:mosaic_iphas}).

@@

\begin{table}
\footnotesize
\resizebox{0.45\textwidth}{!}{
    \begin{tabular}{lccc}
    \toprule
    Name [IPHAS2 ...] & $r$ & $i$ & \ha  \\
    \midrule
J192954.40+181026.1& $17.69\pm0.01$ & $16.12\pm0.01$ & $16.19\pm0.01$ \\
J193011.01+182051.2& $18.55\pm0.02$ & $16.95\pm0.02$ & $17.31\pm0.02$ \\
J193021.52+181954.5& $19.72\pm0.05$ & $17.94\pm0.03$ & $18.47\pm0.04$ \\
J193024.45+181938.3& $19.31\pm0.04$ & $17.55\pm0.02$ & $17.99\pm0.03$ \\
J193033.00+181609.3& $18.25\pm0.01$ & $16.91\pm0.01$ & $16.92\pm0.01$ \\
J193042.48+182317.4& $19.96\pm0.03$ & $18.11\pm0.03$ & $18.48\pm0.03$ \\
    \bottomrule
    \end{tabular}}
    \caption{Candidate \ha-emitters towards Sh 2-82.}
    \label{tbl:emitters}
\end{table}

In previous work, we have shown how the majority of
\ha-emitters seen towards an H{\sc ii} region
are likely to be Classical T Tauri Stars (CTTS),
which are young objects which show \ha\ in emission
due to the presence of hot, infalling gas
which is accreting onto the star
from a circumstellar disk \cite{Barentsen2011a}.
This is likely to be the case in this demonstration
as well. Two of our candidates have recently been identified
as Young Stellar Objects (YSO)
in an investigation of the region by \cite{Yu2012}.
In their study, the authors used 2MASS and Spitzer data
to select likely young stars by selecting objects with
an infrared colour excess, consistent with 
the presence of a heated circumstellar disk.
Although \citeauthor{Yu2012} did not identify
four of our candidates as having
an infrared excess, we note  that each of them
is detected at S/N$>$? in the Spitzer 8.0\,\micron\ image,
which is more likely for a circumstellar-bearing
object than it is for a field star.

Prior to IPHAS, this region was essentially unstudied
at faint magnitudes in visible light.
\cite{Lahulla1985} reported magnitudes for 8 stars in the optical at 
$V < 15$. In contrast, the IPHAS catalogue offers photometry
for 2197 objects in the region down to $r<20$.

Sh 2-82 is one of a wealth of poorly-studied star-forming regions
at low Galactic latitudes,
which have started to become revealed
in recent years through catalogues
of ``bubbles'' seen
in mid-infrared surveys \cite[e.g.][]{Churchwell2006,Simpson2012}
and previously unknown clusters seen 
at near-infrared wavelengths \cite[e.g.][]{Bica2003}.
The optical data provided by IPHAS offers a handle (or lower limit)
on the extinction and distance towards many of these regions,
and can provide information on the optically-unveiled
members of the region, as we demonstrated here.

%TODO: annotate -10 EW text on plot?
%TODO: mention Gaia, e.g. IPHAS for transient identification?
%\subsection{Caveats and lessons learnt}
%\subsection{Comparison with Gaia}?

@@

The observations included in this release
achieved a median seeing of 1.1 arcsec
and a $5\sigma-$depth of $r=21.2\pm 0.5$, $i=20.0\pm 0.3$, \ha$=20.3\pm 0.3$.
The global calibration and photometric repeatability
was found to be accurate at the level of $\sigma=0.03$ mag,
providing a significant improvement over the 
previous data release.
The source catalogue provides the best-available
single-epoch astrometry and photometry
for 219~million unique sources.

In future work, we aim to draw upon the PanSTARRS photometric
survey to further improve the accuracy of our global calibration.
We will also aim to correct the IPHAS photometry for the
radial field distortions,
and aim to produce a catalogue
which details all the secondary detections
to aid time-domain studies.

The data-taking strategy developed for IPHAS
have since been reapplied to carry out a companion INT/WFC survey called UVEX
in U/$g$/$r$/He {\sc i} \citep{Groot2009}
and a southern companion survey in $u$/$g$/$r$/$i$/\ha\ 
called VPHAS+ \citep{Drew2014}.
We intend to use the present data release
as a template for future releases from these companion surveys.

@@
